---
title: "King County Housing Price Analysis"
format: html
bibliography: bibliography.bib
---

## Question 
When purchasing a house in King County, potential buyers often consider factors such as construction quality, size, and age of the property. Using a sample from the King County housing dataset, I explore: **To what extent do these fundamental housing characteristics - construction grade, living area, and year built - explain the variation in housing prices across the county?** 

::: {.callout-note}
Specifically, we are considering:

1. what is the relationship between a house's living area and its price, and does the relationships differ by grade?
2. how have housing prices evolved over time, for different construction grades?
:::

## Intended Audience
This analysis is designed for potential homebuyers and real estate professionals in King County who need to understand the relationship between house characteristics and prices. The information is presented in an accessible way for readers with basic statistical knowledge.

## Dataset

### Data Source

The data used in this analysis comes from the King County House Sales dataset, which is available on [Kaggle](https://www.kaggle.com/code/burhanykiyakoglu/predicting-house-prices/input) [@kaggle_kingcounty]. The dataset contains house sale prices for King County, which includes Seattle and its surrounding areas, with homes sold between May 2014 and May 2015.

### Data Dictionary
::: {.column-margin}
While we focus on these four key features that most directly influence price, the original full dataset contains 21 variables including bathrooms, floors, and waterfront status.
:::


The variables used in our analysis are defined as follows:

| Variable | Description |
|----------|-------------|
| price | Sale price of the house in dollars |
| grade | Grade of the house on King County grading system (1-13)<br>• 1-3: Poor construction<br>• 4-6: Low quality<br>• 7: Average<br>• 8-10: Good quality<br>• 11-13: High quality/Luxury |
| yr_built | Year when the house was built |
| sqft_living | Square footage of living space in the home |

[@kingcounty_grades]

## Data Wrangling
We need to select the only variables we want to explore and pick 200 out of all the observations in the dataset as a sample.
```{r}
#| results: hide
#| warning: FALSE
library(tidyverse) 
library(gridExtra)  
library(viridis)    
library(kableExtra) 
```

```{r}
data=read.csv("kc_house_data.csv")
attach(data)

rows <- sample(1:21613, 200, replace=FALSE)
data_fr <- data[rows,]
data1 <- na.omit(data_fr)
data <- data1 %>%
  group_by(price) %>%
  select(price, grade, yr_built, sqft_living) %>%
  arrange(desc(price))

data
```


## Visualization Analysis

```{r}
# Add log_price to dataset
data$log_price <- log(data$price)
```

::: {.callout-note}
## Data Preprocessing Step
Before creating visualizations, we first transformed the price variable using natural logarithm (log(price)). 
This transformation is common in housing price because:

1. House prices typically follow a right-skewed distribution [@james2013introduction]
2. It helps normalize the data
3. Makes it easier to interpret percentage changes in price
:::


```{r}
#| warning: FALSE
# 1. Enhanced Box Plot with Points - Focus on Grade Distribution

p1 <- ggplot(data, aes(x = factor(grade, levels = sort(unique(grade))), y = log_price)) +
  geom_boxplot(fill = "lightblue", alpha = 0.5) +
  geom_violin(alpha = 0.2) +
  geom_jitter(width = 0.2, alpha = 0.1) +
  theme_minimal() +
  labs(title = "Price Distribution Across Housing Grades in King County",
       subtitle = "Showing price spread and density for each grade level",
       caption = "Graph by Xindi Shan (2024)",
       x = "Grade",
       y = "log(Price)") +
  theme(axis.text.x = element_text(angle = 0))

p1
```

```{r}
#| warning: FALSE
# 2. Faceted Year Analysis - Temporal Patterns by Grade

p2 <- ggplot(data, aes(x = yr_built, y = log_price)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  # Facet by grade
  facet_wrap(~grade, scales = "free_y", ncol = 3) +
  theme_minimal() +
  labs(title = "Price Trends Over Time by Grade",
       subtitle = "Points show individual houses, red line shows linear trend",
       caption = "Graph by Xindi Shan (2024)",
       x = "Year Built",
       y = "log(Price)") +
  theme(strip.background = element_rect(fill = "lightgray"),
        strip.text = element_text(color = "black"))

p2
```

```{r}
# 3. Size-Price Relationship With Grade

p3 <- ggplot(data, aes(x = sqft_living, y = log_price)) +
  # Points colored by grade
  geom_point(aes(color = as.factor(grade)), alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  scale_color_viridis_d(name = "Grade") +
  theme_minimal() +
  labs(title = "Price-Size Relationship by Grade",
       subtitle = "Points colored by grade, black line shows overall trend",
       caption = "Graph by Xindi Shan (2024)",
       x = "Living Area (sqft)",
       y = "log(Price)")

p3
```
::: {.callout-tip}
Note that our analysis focuses on a random sample of 200 houses from the full King County dataset.
:::


## Summary and Interpretation
The examination of King County's housing market reveals several significant patterns. Properties with higher `grade` values consistently command higher `log_price` values. The relationship between `sqft_living` and `log_price` is particularly strong in high-grade properties, reflecting the luxury segment of the Seattle market. Analysis of `yr_built` shows an upward trend in `log_price` across all grades, influenced by the region's tech industry growth. Moreover, larger homes (measured by `sqft_living`) typically correspond to higher `grade` values, indicating a strong relationship between construction quality and house size.

## Function Used
`dplyr` package: `%>%`, `group_by()`, `select()`, `arrange()`, `desc()`

`ggplot2` package: `ggplot()`, `geom_boxplot()`, `geom_violin()`, `geom_jitter()`, `geom_point()`, `geom_smooth()`, `facet_wrap()`, `theme_minimal()`, `labs()`, `scale_color_viridis_d()` 

